---
title: "Google Data Analyst Certificate Capstone Project"
author: "Opeyemi Adesany"
date: "`r Sys.Date()`"
output: html_document
---
##loading the packages that will be needed for the project
Loading the dataset required for analysis

```{r setup, include=FALSE}
#installing library
library(tidyverse)
library(lubridate)
library(readr)
library(ggplot2)
library(readr)
library(hms)
library(geosphere) #geosphere package to compute the distance between two coordinates

```

## Importing the data set into R
After loading the packages that will be needed for the analysis, the next thing is to import the data set

```{r cars}
#importing the data set  
jul_2021 <-read_csv("~/bike_share/202107-divvy-tripdata.csv")

aug_2021 <- read_csv("~/bike_share/202108-divvy-tripdata.csv")

sep_2021 <- read_csv("~/bike_share/202109-divvy-tripdata.csv")

oct_2021 <- read_csv("~/bike_share/202110-divvy-tripdata.csv")

nov_2021 <-read_csv("~/bike_share/202111-divvy-tripdata.csv")

dec_2021 <-read_csv("~/bike_share/202112-divvy-tripdata.csv")

jan_2022 <- read_csv("~/bike_share/202201-divvy-tripdata.csv")

feb_2022 <-read_csv("~/bike_share/202202-divvy-tripdata.csv")

mar_2022 <-read_csv("~/bike_share/202203-divvy-tripdata.csv")

apr_2022 <- read_csv("~/bike_share/202204-divvy-tripdata.csv")

may_2022 <- read_csv("~/bike_share/202205-divvy-tripdata.csv")

jun_2022 <- read_csv("~/bike_share/202206-divvy-tripdata.csv")


```

## Merginging the data sets into single data frame
merging the 12 months of data set into single data frame

```{r}
#merging the last 12 months dataset
merged_12_months_data <- rbind(jul_2021,
                                 aug_2021,
                                 sep_2021,
                                 oct_2021,
                                 nov_2021,
                                 dec_2021,
                                 jan_2022,
                                 feb_2022,
                                 mar_2022,
                                 apr_2022,
                                 may_2022,
                                 jun_2022)

```
inspecting the data frame of the merged data

```{r}
#Inspecting the combined data frame
colnames(merged_12_months_data) #column names in the data frame
nrow(merged_12_months_data) #number of rows in the data frame
dim(merged_12_months_data) #retrieve the dimension of the data frame
head(merged_12_months_data) #
str(merged_12_months_data) #see list of columns and data type
summary(merged_12_months_data) #statistical summary of the data
```
Creating additional column for further analysis

```{r pressure, echo=FALSE}
#creating new column which are ride_duration, ride_distance, week_day_name and month_name to better understand customer usage of the service
merged_12_months_data <- merged_12_months_data %>%
  mutate(ride_duration = as.numeric(difftime(ended_at, started_at,units = "mins")),
         ride_distance = distHaversine(cbind(start_lng, start_lat), cbind(end_lng, end_lat)),
         day_of_week = weekdays(started_at),
         ride_month = months(started_at),
         ride_year = year(started_at))
```

Inspecting the additional column added to the data frame
```{r}
#Inspecting the data frame after addition of column
colnames(merged_12_months_data) #column names in the data frame
nrow(merged_12_months_data) #number of rows in the data frame
dim(merged_12_months_data) #retrieve the dimension of the data frame
head(merged_12_months_data)
str(merged_12_months_data) #see list of columns and data type
summary(merged_12_months_data) #statistical summary of the data
```

Arranging the week day ride and month ride accordily
```{r}
#Arrange the day of the week and month accordingly
#order for day of the week
merged_12_months_data$day_of_week <- ordered(merged_12_months_data$day_of_week,
                                               levels = c("Sunday", 
                                                          "Monday", 
                                                          "Tuesday",
                                                          "Wednesday", 
                                                          "Thursday",
                                                          "Friday",
                                                          "Saturday"))

#order for month
merged_12_months_data$ride_month <- ordered(merged_12_months_data$ride_month, 
                                            level = c("July",
                                                      "August",
                                                      "September",
                                                      "October",
                                                      "November",
                                                      "December",
                                                      "January",
                                                      "February",
                                                      "March", 
                                                      "April",
                                                      "May",
                                                      "June"))
```

Inspecting the arranged day and month of ride accordily
```{r}
#inspecting the data frame
str(merged_12_months_data)
colnames(merged_12_months_data)
head(merged_12_months_data)
```
Removing unwanted column from the data frame

```{r}
#removing unwanted column from the data set
merged_12_months_data <- merged_12_months_data %>%
  select(-c(start_lat:end_lng))
```

Inspecting the new data frame after removal of unwanted column

```{r}
str(merged_12_months_data)
colnames(merged_12_months_data)
head(merged_12_months_data)
```
Further cleaning of the data set to remove trips not start at a station and not ending at a station and also ride duration less than 5 mins

```{r}
#further cleaning of the data set to remove trips not start at a station and not ending at a station and also ride duration less than 5 mins
cleaned_12_months_data <- merged_12_months_data %>%
  filter(!is.na(start_station_name)) %>%
  filter(!is.na(end_station_name)) %>%
  filter(ride_duration >= 5) %>%
  filter(ride_distance > 0)

```

Inspect the data frame after removing some variables

```{r}
#Inspecting the data frame after further cleaning
colnames(merged_12_months_data) #column names in the data frame
nrow(merged_12_months_data) #number of rows in the data frame
dim(merged_12_months_data) #retrieve the dimension of the data frame
head(merged_12_months_data)
str(merged_12_months_data) #see list of columns and data type
summary(merged_12_months_data) #statistical summary of the data

```

Running preliminary statistics to know the nature of the data frame

```{r}
#Doing preliminary statistic analysis to check for data distribution
count_station <- cleaned_12_months_data %>%
count(start_station_name) %>%
rename(count_station = n) %>%
arrange(desc(count_station))
head(count_station)
tibble(count_station)
```

```{r}
#summary of ride by each casual and member
member_category_summary <- cleaned_12_months_data %>%
  group_by(member_casual) %>%
  summarise(total_ride = n(),
            average_ride = as.numeric(mean(ride_duration)),
            maximum_ride= as.numeric(max(ride_duration)),
            minimum_ride = as.numeric(min(ride_duration)),
            average_distance = as.numeric(mean(ride_distance)),
            maximum_distance = as.numeric(max(ride_distance)),
            minimum_distance = as.numeric(min(ride_distance))
  ) %>%
  ungroup()
tibble(member_category_summary)
```

```{r}
#summary of ride by member and casual on weekdays
summary_ride_member_casual <- cleaned_12_months_data %>%
  group_by(member_casual, day_of_week, ride_month,ride_year) %>%
  summarise(total_ride = n(),average_ride = as.numeric(mean(ride_duration))
  )%>%
  ungroup()

head(summary_ride_member_casual)
```

```{r}
#we want to count number of ride for each month
summary_ride_by_month <- cleaned_12_months_data %>%
  count(ride_month)
  
head(summary_ride_by_month)

```

Analysing the data to get some insight from the data using group_by() and summarise() to aggregate valuse from column of interest.
I extracted membership type, rideable type, ride year, ride month, day of week. 
```{r}
result_12_months_data <- cleaned_12_months_data %>%
  group_by(member_casual, rideable_type, ride_year, ride_month, day_of_week) %>%
  summarise(number_of_rides = n(), avg_ride_duration = mean(ride_duration), avg_ride_distance = mean(ride_distance))

tibble(result_12_months_data)
dim(result_12_months_data)
tibble(result_12_months_data)
```

Exporting the insight made from the analysis
```{r}
#exporting my code into csv file
result_file_result_12_months_data <- paste(getwd(),"/result_12_months_data.csv", sep = "")
write_csv(result_12_months_data, result_file_result_12_months_data)

result_file_summary_ride_by_month <- paste(getwd(),"/summary_ride_by_month.csv", sep = "")
write_csv(summary_ride_by_month, result_file_summary_ride_by_month)


result_file_summary_ride_member_casual <- paste(getwd(),"/summary_ride_member_casual.csv", sep = "")
write_csv(summary_ride_member_casual, result_file_summary_ride_member_casual)

result_file_member_category_summary  <- paste(getwd(),"/member_category_summary .csv", sep = "")
write_csv(member_category_summary , result_file_member_category_summary )

```

Thank you.